### 2020-05-19 manuscript.tex
adding simulated data parts, hnn discussion, edits to cryptic species discussion

To best analyze the cryptic diversity within \textit{Ceratopteris} we gathered samples from across its pan-tropical distribution. We utilized RADSeq to generate a very large genomic dataset, which was processed using the ipyrad pipeline \autocite{Eaton2020} for downstream analysis. Our three-part approach for assessing cryptic diversity takes advantage of the flexibility of RADSeq data, and provides a unique window into the processes of speciation and diversification in a complex genus.

@@

We started with 90 samples, 56 from herbarium specimens and 34 from silica-dried tissue. These samples represented five of seven named species of \textit{Ceratopteris}, and one sample of \textit{Acrostichum aureum} from Australia for use as an outgroup. The two missing species of \textit{Ceratopteris} were cryptic species of \textit{C. thalictroides} (L.) Brongn. \autocite{Masuyama2010}: \textit{C. oblongiloba} Masuyama \& Watano, known from Southeast Asia; and \textit{C. froesii} Brade, endemic to Brazil. Both have relatively narrow ranges and few herbarium species, especially \textit{C. froesii}. We were unable to obtain any material from herbarium specimens labeled as either of these species, and none of the plants collected in the field keyed out to \textit{C. oblongiloba} or \textit{C. froesii}.

@@

@@

Unfortunatley, all samples of \textit{Ceratopteris richardii} either failed in sequencing, or had very poor sequencing quality. After an initial examination of the results of these samples, we chose to remove them from our final analyses. We did not want to attempt to make any inferences about this important species with poor-quality data.

@@

 %Samples of \textit{C. richardii} were also found in multiple places in the phylogeny; however, due to their lower data quality this is not surprising and little inference can be made about its phylogenetic position.

@@

\textit{Ceratopteris cornuta}, \textit{C. pteridoides}, and \textit{C. richardii} are known as a diploid (n = 39) \autocite{Adjie2007, Hickok1977}; \textit{Ceratopteris thalictroides} and \textit{C. gaudichaudii} are known as tetraploid, with some cytotypic variation (n = 77, 78) \autocite{Adjie2007, Masuyama2010}.

@@

@@

@@

New World \textit{C. thalictroides} is strongly differentiated from Old World \textit{C. thalictroides} in our population structure analysis, and shows virtually no gene flow between any other species at \textit{K} greater than 3. This strong distinction between Old and New World is present in the phylogeny as well: New World \textit{C. thalictroides} is placed in a clade as sister to \textit{C. gaudichaudii}, and this clade is in turn sister to Old World \textit{C. thalictroides} (Fig. \ref{phy}.

@@

All of the cryptic species in the \textit{Ceratopteris thalictroides} complex are known to be tetraploid (\textit{n} = 77, 78; \cite{Masuyama2010}). Several authors have proposed that different cryptic species have been derived via differing mechanisms of polyploidy (i.e., allo- vs. autopolyploid). \textcite{Adjie2007} identified Asian \textit{C. thalictroides} \textit{sensu latu} as paraphyletic and hypothesized that \textit{C. thalictroides sensu strictu} and the cryptic species \textit{C. gaudichaudii} and \textit{C. oblongiloba} are allopolyploid hybrids between \textit{C. cornuta} and unnamed, potentially extinct, diploid progenitors. \textcite{McGrath1994} suggested that New World \textit{C. thalictroides} could either be an autopolyploid, or an allopolyploid hybrid between \textit{C. ricahrdii} and another diploid species.

@@

Weak reproductive boundaries are a theme across ferns, evidenced by numerous reticulate species complexes (\emph{e.g.}, \cite{Barrington1989,  Paris1989, Sessa2012a, Yatabe2009}), and intergeneric hybridization across 60 MY of evolution \autocite{Rothfels2015}. Although ferns do have some pre- and post-zygotic barriers \autocite{Haufler2016}, they are not as strong as in angiosperms (e.g., \cite{De_Nettancourt1997, Lafon-Placette2016}). In addition, fern spores are very easily dispersed \autocite{Tryon1970, Smith1972}, which limits isolation by distance.

In the case of \textit{Ceratopteris}, its large range may be influenced by the Intertropical Convergence Zone (ITCZ) (see \cite{Dettmann1992, LloydTax1974, Schneider2014}). The ITCZ is created by the rising portion of the Hadley Cell circulation in both hemispheres; it moves slightly above and below the equator with the seasons, following the thermal equator \cite{Schneider2014}. It has been hypothesized as a possible mechanism of transport of Asian ferns to Hawaii \autocite{Geiger2007}, and could carry spores great distances from one area of the tropics to another, provided they make it high enough into the atmosphere, perhaps by monsoons or hurricanes. Continuous dispersal and/or movement of propagules is recognized as an important factor in widespread angiosperm hydrophytes \autocite{Les2003, Spalik2014}. The ITCZ could be a means of dispersal for \textit{Ceratopteris}, which would help explain the continued gene flow across the large range of its species such as Old World \textit{C. thalictroides}, or \textit{C. cornuta}. 
%All of these factors may be limiting genetic diversification and enabling hybridization in \textit{Ceratopteris}, despite its pan-tropical distribution. Continued gene flow may be contributing to the morphological similarities in the group, but there are certainly other factors at play.

Even with a high level of dispersal, there is potential evidence for sympatric speciation in \textit{Ceratopteris}. Two specimens collected meters apart in Australia keyed to \textit{C. thalictroides sensu strictu} and \textit{C. gaudichaudii} var. \textit{vulgaris}. In addition, specimens of Old World \textit{C. thalictroides} and New World \textit{C. thalictroides} were obtained from South America. [other examples: \textit{Asplenium nidus}, ...]

%crypsis is known in other groups of ferns
%pteridium
%polypodium
%asplenium
%other animals and plants

One of the most confusing aspects of \textit{Ceratopteris} is the disparity between morphological and genetic distinction between species. The three diploid species are relatively morphologically distinct \autocite{LloydTax1974}, especially in comparison to the cryptic species complex of \textit{C. thalictroides sensu latu}. However, all the diploid species in the genus are known to hybridize with one another \autocite{hickok1974, Hickok1977, LloydTax1974}. In contrast, \textit{C. thalictroides} and associated cryptic species are much more similar morphologically, yet different populations (some named as cryptic species) can be almost entirely reproductively isolated \autocite{Hickok1979, Masuyama2010}.

A potential explanation for morphological stasis in \textit{Ceratopteris} relies on the influence ecological pressures \autocite{Bickford2007}. For a fern, \textit{Ceratopteris} has an unusually short life cycle of only about four months from spore to spore-bearing adult \autocite{Stein1971}. It grows in temporary water sources such as pond edges, swamps, or tarot patches \autocite{LloydTax1974}. Because of its ephemeral habitat, there are likely selective pressures for maintaining morphological stasis and rapid generation times. Fossil spore and leaf evidence also support morphological stasis in the Ceratopteridoideae dating back about 47 MY \autocite{Dettmann1992, Rozefelds2016-dr}. Fossil leaf impressions of the extinct genus \textit{Tecaropteris} from Australia are very similar morphologically to extant \textit{Ceratopteris} \autocite{Rozefelds2016-dr}. Australia has become increasingly dry and seasonal since the Eocene, when \textit{Tecaropteris} lived \autocite{Rozefelds2016-dr, McKenna2010}. Today, \textit{Ceratopteris} is exceedingly difficult to find during the dry season in northeastern Australia, but plentiful during the wet season, even growing as a weed in garden ponds (Dr. Ashley Field, personal communication). Almost all leaves of \textit{Ceratopteris} species are very thin, delicate, and herbaceous \autocite{LloydTax1974}. When removed from the plant they wilt into an unrecognizable state within minutes (personal observation, SPK). These low-energy investment leaves can be grown quickly and without an excess of resources from the plant \autocite{Reich2014, Wright2004}, potentially as a result of the fast life cycle employed by \textit{Ceratopteris}; any major shift in leaf development might inhibit \textit{Ceratopteris} from proceeding through its life cycle before its water source dries up. Therefore, despite the genetic diversity in the genus, morphological diversity has changed very little due to niche constraints.

@@

>>>>

To best analyze the cryptic diversity within \textit{Ceratopteris} we gathered samples from across its pan-tropical distribution. We utilized RADSeq to generate a large genomic dataset, which was processed using the ipyrad pipeline \autocite{Eaton2020} for downstream analysis. Our three-part approach for assessing cryptic diversity takes advantage of the flexibility of RADSeq data, and provides a unique window into the processes of speciation and diversification in a complex genus.

@@

We started with 90 samples, 56 from herbarium specimens and 34 from silica-dried tissue. These samples represented five of seven named species of \textit{Ceratopteris}, and one sample of \textit{Acrostichum aureum} from Australia for use as an outgroup. We also included a sample of the common lab strain of \textit{C. richardii} (\textit{Hnn}). The two missing species of \textit{Ceratopteris} were cryptic species of \textit{C. thalictroides} (L.) Brongn. \autocite{Masuyama2010}: \textit{C. oblongiloba} Masuyama \& Watano, known from Southeast Asia; and \textit{C. froesii} Brade, endemic to Brazil. Both have relatively narrow ranges and few herbarium species, especially \textit{C. froesii}. We were unable to obtain any material from herbarium specimens labeled as either of these species, and none of the plants collected in the field keyed out to \textit{C. oblongiloba} or \textit{C. froesii}.

@@

\textbf{2.8. \textit{In silico} analysis of the \textit{Ceratopteris richardii} genome}

Our sampling included relatively few individuals of \textit{Ceratopteris richardii}, so we utilized the published genome from \textcite{Marchant2019} to perform an \textit{in silico} RADseq digest, creating a pseudo-indiviual to include in our analyses. We used the program ddRADSeqTools \autocite{Mora-Marquez2017} to digest the genome with the same enzymes used in our \textit{in vivo} digest (PstI and BfaI). Then we performed a simulated ddRADSeq run with paired-end reads, demultiplexing, and trimming using ddRADSeqTools. This simulated indidivuals was added to our other samples and processed via the ipyrad pipeline. Finally, we ran STRUCTURE for this set of individuals.

@@

Unfortunately, all samples of \textit{Ceratopteris richardii} (including \textit{Hnn}) either failed in sequencing, or had very poor sequencing quality. After an initial examination of the results of these samples, we chose to remove them from our final analyses. We did not want to attempt to make any inferences about this important species, nor the lab strain, with poor-quality data.

@@

@@

\textit{Ceratopteris cornuta}, \textit{C. pteridoides}, and \textit{C. richardii} are known as a diploid (\textit{n} = 39) \autocite{Adjie2007, Hickok1977}; \textit{Ceratopteris thalictroides} and \textit{C. gaudichaudii} are known as tetraploid, with some cytotypic variation (\textit{n} = 77, 78) \autocite{Adjie2007, Masuyama2010}. Our ploidy analysis used allelic ratios to determine the ploidy of samples. Diploids would have a 1:1 allele ratio in accordance with the two parental genomes present. Triploids (derived via hybridization) would have a 2:1 allele ratio, with two sets of chromosomes from a unreduced spore, and another from a haploid spore. Tetraploids could either have a 1:1 or 3:1 allele ratio: the former would be representative of an autopolyploid derived from a genome double event, or an allopolyploid with two sets of homeologous chromosomes; the latter would be representation of an allopolyploid that has undergone genomic restructuring  (i.e. compensated aneuploidy, \cite{Sigel2016}).

@@

\textbf{3.4. Analysis of simulated RADSeq data}

We attempted to create a pseudo-individual to including in our analysis using the \textit{Ceratopteris richardii} genome assembly. We were able to obtain 9,047,554 simulated raw sequences utilizing an \textit{in silico} RADSeq digest. These reads, in combination with the rest of our \textit{in vivo} data, were processed with the ipyrad pipeline. The ipyrad pipeline consists of 7 steps, each performing one aspect of RADSeq data processsing \autocite{Eaton2020}. Our pseduo-individual of \textit{C. richardii} performed nearly identically to all of the other samples until Step 6. Here, similar sequences are clustered across samples, and then these shared sequences are mapped to a reference (our reference was a \textit{de novo} assembly, because the quality and coverage of the \textit{C. richardii} genome was too poor to serve as a reference). This requires that all individuals are related relatively close to one another, so that they share a large number of sequences. Very few sequences were retained for our pseudo-individual of \textit{C. richardii}, indicating that the sequences obtained are too different from the \textit{in vivo} samples. In order to retain enough SNPs to run STRUCTURE, we had to reduce the minimum number of individuals to have data at a particular locus to just 3. In our entirely \textit{in vivo} dataset, this minimum number of individuals was 30; we designated such a high number to reduce missing data. In our dataset with the pseud-individual of \textit{C. richardii}, a vast majority of our retained loci had missing data.

The differences between the pseudo-individual derived from the \textit{Ceratopteris richardii} and our other samples is mostly likely due to the fact that the current assembly (v. 1.1) only represents about one third of the total genome, and consists of mostly short reads. Long reads that can span the many repetitive elements in the genome comprise only 0.03\% of the final assembly \autocite{Marchant2019}. Because of these factors, the current genome assembly contains small amount of short sequences that are difficult to accurately compare to the fragments generated by \textit{in vivo} RADSeq. We have included the results from our \textit{in silico} RADSeq digest in Appendix A: Supplementary data 2, but do not present them in the main manuscript.

@@

Not included in our results are specimens of the model fern \textit{Ceratopteris richardii}, including the lab strain \textit{Hnn}. All individuals samples we had of this species yielded very poor sequencing results. We felt it was appropriate to remove these samples because we did not want to try and infer anything about this important species with poor quality data. Future work will be aimed at including wild collections of \textit{C. richardii} as well as the lab strain \textit{Hnn} to determine the relationship of this species to the rest of the genus.

@@

New World \textit{C. thalictroides} is strongly differentiated from Old World \textit{C. thalictroides} in our population structure analysis, and shows virtually no gene flow between any other species at \textit{K} greater than 3. This strong distinction between Old and New World is present in the phylogeny as well: New World \textit{C. thalictroides} is placed in a clade as sister to \textit{C. gaudichaudii}, and this clade is in turn sister to Old World \textit{C. thalictroides} (Fig. \ref{phy}). \textit{Ceratopteris gaudichaudii} has been hypothesized to be an allopolyploid hybrid \autocite{Adjie2007}; perhaps the derivation of its sister species, New World \textit{C. thalictroides}, could have been from a hybridization event with one of the parents of \textit{C. gaudichaudii}, or genome double to form an autotetraploid, followed by long distance dispersal.

@@

All of the cryptic species in the \textit{Ceratopteris thalictroides} complex are known to be tetraploid (\textit{n} = 77, 78; \cite{Masuyama2010}). Several authors have proposed that different cryptic species have been derived via differing mechanisms of polyploidy (i.e., allo- vs. autopolyploid). \textcite{Adjie2007} identified Asian \textit{C. thalictroides} \textit{sensu latu} as paraphyletic and hypothesized that \textit{C. thalictroides sensu strictu} and the cryptic species \textit{C. gaudichaudii} and \textit{C. oblongiloba} are allopolyploid hybrids between \textit{C. cornuta} and unnamed, potentially extinct, diploid progenitors. \textcite{McGrath1994} suggested that New World \textit{C. thalictroides} could either be an autopolyploid, or an allopolyploid hybrid between \textit{C. richardii} and another diploid species.

@@

\textbf{4.3.1. Biogeography and reproductive boundaries}

Weak reproductive boundaries are a theme across ferns, evidenced by numerous hybrids and reticulate species complexes (\emph{e.g.}, \cite{Barrington1989,  Paris1989, Sessa2012a, Sigel2016, Yatabe2009}), and intergeneric hybridization across 60 MY of evolution \autocite{Rothfels2015}. Although ferns do have some pre- and post-zygotic barriers \autocite{Haufler2016}, they are not as strong as in angiosperms (e.g., \cite{De_Nettancourt1997, Lafon-Placette2016}). In addition, fern spores are very easily dispersed \autocite{Tryon1970, Smith1972}, which limits isolation by distance.

In the case of \textit{Ceratopteris}, its large range may be influenced by the Intertropical Convergence Zone (ITCZ) (see \cite{Dettmann1992, LloydTax1974, Schneider2014}). The ITCZ is created by the rising portion of the Hadley Cell circulation in both hemispheres; it moves with the seasons, following the thermal equator \cite{Schneider2014}. It has been hypothesized as a possible mechanism of transport of Asian ferns to Hawaii \autocite{Geiger2007}, and could carry spores great distances from one area of the tropics to another, provided they make it high enough into the atmosphere, perhaps by monsoons or hurricanes. Continuous dispersal and/or movement of propagules is recognized as an important factor in widespread angiosperm hydrophytes \autocite{Les2003, Spalik2014}. The ITCZ could be a means of dispersal for \textit{Ceratopteris}, which would help explain the continued gene flow across the large range of its species such as Old World \textit{C. thalictroides}, or \textit{C. cornuta}. 
%All of these factors may be limiting genetic diversification and enabling hybridization in \textit{Ceratopteris}, despite its pan-tropical distribution. Continued gene flow may be contributing to the morphological similarities in the group, but there are certainly other factors at play.

Even with a high level of dispersal, there is potential evidence for reproductive boundaries and sympatric speciation in \textit{Ceratopteris}. Two specimens collected meters apart in Australia keyed to \textit{C. thalictroides sensu strictu} and \textit{C. gaudichaudii} var. \textit{vulgaris}. In addition, specimens of Old World \textit{C. thalictroides} and New World \textit{C. thalictroides} were obtained from similar regions in South America. Sympatric cryptic species are known from other genera of ferns \autocite{Patel2019, Yatabe2009}, lichens \autocite{Del_Carmen_Molina2011}, angiosperms \autocite{Les2015, Soltis2007}, as well as in animals \autocite{Amor2014, Hebert2004, Nygren2014}. How such a readily dispersable fern such as \textit{Ceratopteris} can diversify sympatrically is perplexing. One mechanism of differentiation could be minor ecologically separation. All \textit{Ceratopteris} species occupy very similar habitats \autocite{LloydTax1974, Masuyama2010}. However, in the case of the two Australian specimens collected at the same locality, the individual of Old World \textit{C. thalictroides} was growing in full sun, while \textit{C. gaudichaudii} was growing in deep shade. How reproductive boundaries have evolved between these co-occurring cryptic species area of future work.

%They can be a monophyletic species complex, very recently derived from a common ancestor and in the process of fully speciating \cite{Del_Carmen_Molina2011, Hebert2004}. Alternatively, cryptic species can be paraphyletic and separated by considerable evolutionary time, yet look similar and occupy comparable niches \cite{Amor2014, Cunnington2005}.

\textbf{4.3.2. Morphology and ecology}

One of the most confusing aspects of \textit{Ceratopteris} is the disparity between morphological and genetic distinction between species. The three diploid species are relatively morphologically distinct \autocite{LloydTax1974}, especially in comparison to the cryptic species complex of \textit{C. thalictroides sensu latu}. However, all the diploid species in the genus are known to hybridize with one another \autocite{hickok1974, Hickok1977, LloydTax1974}. In contrast, \textit{C. thalictroides} and associated cryptic species are much more similar morphologically, yet different populations (some named as cryptic species) can be almost entirely reproductively isolated \autocite{Hickok1979, Masuyama2002, Masuyama2010}. However, Throughout the genus, however, the body plan, life history, and niche requirements of all species are quite similar, indicating limited ecological divergence.

A potential explanation for morphological stasis in \textit{Ceratopteris} relies on the influence of ecological pressures \autocite{Bickford2007}. For a fern, \textit{Ceratopteris} has an unusually short life cycle of only about four months from spore to spore-bearing adult \autocite{Stein1971}. It grows in temporary water sources such as pond edges, swamps, or tarot patches \autocite{LloydTax1974}. Because of its ephemeral habitat, there are likely selective pressures for maintaining morphological stasis and rapid generation times. Fossil spore and leaf evidence also support morphological stasis in the Ceratopteridoideae dating back about 47 MY \autocite{Dettmann1992, Rozefelds2016-dr}. Fossil leaf impressions of the extinct genus \textit{Tecaropteris} from Australia are very similar morphologically to extant \textit{Ceratopteris} \autocite{Rozefelds2016-dr}. Australia has become increasingly dry and seasonal since the Eocene, when \textit{Tecaropteris} lived \autocite{Rozefelds2016-dr, McKenna2010}. Today, \textit{Ceratopteris} is exceedingly difficult to find during the dry season in northeastern Australia, but plentiful during the wet season, even growing as a weed in garden ponds (Dr. Ashley Field, personal communication). Almost all leaves of \textit{Ceratopteris} species are very thin, delicate, and herbaceous \autocite{LloydTax1974}. When removed from the plant they wilt into an unrecognizable state within minutes (SPK, personal observation). These low-energy investment leaves can be grown quickly and without an excess of resources from the plant \autocite{Reich2014, Wright2004}, potentially as a result of the fast life cycle employed by \textit{Ceratopteris}; any major shift in leaf development might inhibit \textit{Ceratopteris} from proceeding through its life cycle before its water source dries up. Therefore, despite the genetic diversity in the genus, morphological diversity has changed very little due to niche constraints.

@@

Supplementary data 2: Simulated data
